---
title: "Exercise 3 - HCA & Heatmaps"
author: "Nilima Ajaikumar, Hagay Jalon, and Gian Zlupko"
date: '2022-10-24'
output: html_document
---


# Exercise 3
## Hierarchical Cluster Analysis and Heatmaps

## Part I: 
#### Please replicate the code and markdown provided in the readings (Bowers) for the mtcars dataset to visualize it using hierarchical cluster analysis heatmaps following the recommendations from the readings, including using uncentered correlation.


Libraries used throughout exercise 
```{r, warning = FALSE, message = FALSE }
library(ggplot2)
library(tidyverse) 
library(broom)
library(data.table) 
library(ComplexHeatmap)
library(circlize)
library(hopach)
```

Load data 
```{r}
# use first seven variables in mtcars data set 
data <- mtcars[, c(1:7)]

# scale the data set 
data.scale<-scale(data)

# inspect the data set 
head(data.scale)
```

Generate initial heatmaps 
```{r}
# heatmap with no clustering
Heatmap(data.scale, name = "mtcars heatmap", 
        cluster_rows=FALSE , cluster_columns = FALSE)

# now cluster by rows but not columns 
Heatmap(data.scale, name = "mtcars heatmap", 
        cluster_columns = FALSE)
```


```{r}
# cluster by rows and columns 
Heatmap(data.scale, name = "mtcars heatmap")
```

Generate custom function to calculate uncentered correlation

The function creates a distance matrix, which is accessed in the `hopach` library. Per the package documentation, distance matrices contain pair wise distances between all rows of the matrix.

```{r}
uncenter.dist<-function(m) {
  as.dist(as.matrix(distancematrix(m, d="cosangle")))
}

```



```{r}
# apply function to create distance matrix on the data set
row.clus<-hclust(uncenter.dist(data.scale), method = "ave")
col.clus<-hclust(uncenter.dist(t(data.scale)), method = "ave")

# apply uncentered correlation as clustering technique and average linkage as agglomeration method 
Heatmap(data.scale, name = "mtcars heatmap", 
        cluster_rows=row.clus, cluster_columns=col.clus)

```


Heatmaps with dichotomous variables 
```{r}

ht_inline<-Heatmap(mtcars[, c(8)], name = "Cylinder Config",
                   col = colorRamp2(c(0,1),c("white", "black")),
                   heatmap_legend_param = list(at = c(0,1),
                         labels = c("V-shaped", "Inline")),
                   width = unit(0.5,"cm"))

ht_manual<-Heatmap(mtcars[, c(9)], name = "Transmission",
                   col = colorRamp2(c(0,1),c("white", "black")),
                   heatmap_legend_param = list(at = c(0,1),
                         labels = c("Automatic", "Manual")),
                   width = unit(0.5,"cm"))
```


```{r}
ht_main = Heatmap(data.scale, name = "cluster rows",
                  cluster_rows=row.clus, cluster_columns = col.clus)

# draw() command used to put together the heatmap and the annotation 
draw(ht_main+ht_manual+ht_inline, auto_adjust = FALSE) 
```

Same HCA heatmap as above but without row labels: 
```{r}
draw(ht_main+ht_manual+ht_inline) 
```


## Part II: 

#### Select at least three different options/alternatives from ComplexHeatmap's first three chapters, and apply them to the mtcars dataset. For example, such as using different clustering algorithms (k-means, etc.), separation of heatmaps, stacking, different ways to annotate (columns and rows and different graphing options beyond black/white blocks), etc. Please provide some text in your markdown discussing what options/alternatives you selected, why this was interesting, and any challenges you experienced with running the visualizations. Provide a few sentences to interpret each new heatmap visualization.


mtcars data set 

```{r}
# create new data set 
data_mat <- as.matrix(mtcars) 

# scale data 
data_mat <- scale(data_mat) 
```

Custom HCA heatmap 1: 

Visualize a heatmap using a categorical variable for annotation with more than two levels. 

```{r}
# update row and col clustering with new data set input into the custom functions created 
row.clus<-hclust(uncenter.dist(ht_1_dat), method = "ave")
col.clus<-hclust(uncenter.dist(t(ht_1_dat)), method = "ave")

# remove the cylinder variable from the data matrix; will use cylinders as annotation 
ht_1_dat <- data_mat[, -2]

# set cyl as annotation on heatmap 
ht_cyl <-Heatmap(mtcars[, c(2)], name = "Cyl Count",
                   col = colorRamp2(c(4,6,8),c("light grey", "grey", "black")),
                   heatmap_legend_param = list(at = c(0,1,2),
                         labels = c("4", "6", "8")),
                   width = unit(0.5,"cm")) 

ht_main <- Heatmap(ht_1_dat, name = "cluster rows", 
                   cluster_rows = row.clus, cluster_columns = col.clus)

# put together annotated heatmap using draw() function 
draw(ht_main + ht_cyl, auto_adjust = FALSE) 

```

Heatmap description: For the HCA heatmap above, we used uncentered correlation as our clustering technique and average linkage as our agglomeration method. Our primary focus was to produce a heatmap that would allow us to compare cars with 4, 6, and 8 cylinder engines. To do so, we visualized the cylinder variable from the mtcars data set as an annotation for the heatmap. 


Heatmap interpretation: The cylinder count for car engines can tell us a lot about what to expect from a car. More cylinders often means more horsepower and less MPG. The heatmap confirms these assumptions while giving more patterns to interpret. First, we see that 8 cylinder engines also tend to be heavier than cars with fewer cylinders. In addition, we also see that most of the 8 cylinder cars have automatic transmissions whereas the cars with fewer cylinders have a greater balance of manual and automatic transmissions. Thus, the use of an HCA heatmap was interesting for our interest in cylinder count. It allowed us to confirm some of our assumptions about the general relationships between cylinder count, horsepower, and MPG while also providing a much richer and nuanced assessment of key differences between cars with different cylinder counts. 


Challenges: For the HCA heatmap above, we visualized engine cylinder count as a categorical variable for annotation. We ran into challenges creating an categorical annotation with more than two levels as was shown in the given code in course readings. Specifically, during our early attempts, we received the error: "The length of 'at' should be the same as the length of 'labels'. Initially, this error message appeared somewhat cryptic to us. However, upon looking at the code chunk used to edit the data corresponding to heatmap annotation, we realized that we needed to pass a third value through the 'at' command in the that the length of our variable needed to match the `heatmap_legend_param` argument. After adding a third value, corresponding with the three levels of our categorical variable, we were then able to successfully render the visualization.


Custom HCA heatmap 2: 

k-means clustering on rows and columns Manhattan distance 

and use of post_fun 


```{r}

# first heatmap using k-means clustering 
ht1_kmeans <- Heatmap(matrix = data_mat, 
        row_km = TRUE, column_km = TRUE, 
        row_title = "k-means clustering")



# second heatmap using uncenetered correlation as clustering technique 

ht2_uncencor <-  Heatmap(matrix = data_mat, 
                        cluster_rows = row.clus, cluster_columns = col.clus, 
                        row_title = "uncentered correlation clustering") 


# create a single vector matrix to store the car titles n  

# combine heat maps into heatmap list object and render output with space between 
ht_list1 <- ht1_kmeans + ht2_uncencor
draw(ht_list1, ht_gap = unit(.5, "cm"))


```


Heatmap description: 


Heatmap Interpretation: 


Challenges: 







Custom HCA heatmap 3: 
Comparison of U.S. and International car manufacturers. New variable created from the car model names to determine whether the car was made by U.S. manufacturers or by international car makers. Weight seemed to be a key difference. Three U.S. cars were appeared to be substantially more heavy than international counterparts. 
 
```{r}
library(xlsx) 
#write.xlsx(mtcars,"mtcars.xlsx") 
# read back into R with first column as true to get a character vector; this export-import step allowed us to grab the column for the car names, which we could not index originally. 
mt_full <- read.xlsx("mtcars.xlsx", sheetIndex = 1) 

# rename the car model name column and store as character 
mt_full <- mt_full %>% 
  rename(model = NA.) %>% 
  mutate(model = as.character(model)) %>% 
  mutate(international = ifelse(model == "Cadillac Fleetwood" | model == "Lincoln Continental" | model == "Chrysler Imperial" | model == "Dodge Challenger" | 
                                 model == "Camarro Z28" | model == "Pontiac Firebird" | model == "Ford Pantera L", 0, 1)) 

# remove the international label for now; will use the international categorical variable to annotate the heatmap
mt_model <- mt_full %>% 
  select(-c(international, model)) 

# convert data frame to matrix and scale the matrix 
mt_model <- as.matrix(mt_model) 
mt_model <- scale(mt_model) 

# add histogram boxplot for weight of cars 
wt_mat <- mt_model[,6] # create matrix subset for wt variable 
histo_anno = rowAnnotation(Weight = anno_barplot(mt_wt, gp = gpar(fill = 3))) 

# format the international variable for
ht_intl <-Heatmap(mt_full[, 13], name = "Int'l Model", right_annotation = histo_anno,
                   col = colorRamp2(c(0,1),c("grey", "black")),
                   heatmap_legend_param = list(at = c(0,1),
                         labels = c("No", "Yes")),
                   width = unit(0.5,"cm"), column_title = "U.S. and International Cars")

ht_models <- Heatmap(mt_model, name = "cluster rows", 
                   cluster_rows = row.clus, cluster_columns = col.clus)

# combine the heatmap with the annotation 
ht_list3 <- ht_models + ht_intl
draw(ht_list3, ht_gap = unit(.5, "cm"))

```
 






