---
title: "Exercise 3 - HCA & Heatmaps"
author: "Nilima Ajaikumar, Hagay Jalon, and Gian Zlupko"
date: '2022-10-24'
output: html_document
---


# Exercise 3
## Hierarchical Cluster Analysis and Heatmaps

## Part I: 
#### Please replicate the code and markdown provided in the readings (Bowers) for the mtcars dataset to visualize it using hierarchical cluster analysis heatmaps following the recommendations from the readings, including using uncentered correlation.


Libraries used throughout exercise 
```{r, warning = FALSE, message = FALSE }
library(ggplot2)
library(tidyverse) 
library(broom)
library(data.table) 
library(ComplexHeatmap)
library(circlize)
library(hopach)
```

Load data 
```{r}
# use first seven variables in mtcars data set 
data<-mtcars[, c(1:7)]

# scale the data set 
data.scale<-scale(data)

# inspect the data set 
head(data.scale)
```

Generate initial heatmaps 
```{r}
# heatmap with no clustering
Heatmap(data.scale, name = "mtcars heatmap", 
        cluster_rows=FALSE , cluster_columns = FALSE)

# now cluster by rows but not columns 
Heatmap(data.scale, name = "mtcars heatmap", 
        cluster_columns = FALSE)
```


```{r}
# cluster by rows and columns 
Heatmap(data.scale, name = "mtcars heatmap")
```

Generate custom function to calculate uncentered correlation

The function creates a distance matrix, which is accessed in the `hopach` library. Per the package documentation, distance matrices contain pair wise distances between all rows of the matrix.

```{r}
uncenter.dist<-function(m) {
  as.dist(as.matrix(distancematrix(m, d="cosangle")))
}

?distancematrix
```



```{r}
# apply function to create distance matrix on the data set
row.clus<-hclust(uncenter.dist(data.scale), method = "ave")
col.clus<-hclust(uncenter.dist(t(data.scale)), method = "ave")

# apply uncentered correlation as clustering technique and average linkage as agglomeration method 
Heatmap(data.scale, name = "mtcars heatmap", 
        cluster_rows=row.clus, cluster_columns=col.clus)


```


Heatmaps with dichotomous variables 
```{r}

ht_inline<-Heatmap(mtcars[, c(8)], name = "Cylinder Config",
                   col = colorRamp2(c(0,1),c("white", "black")),
                   heatmap_legend_param = list(at = c(0,1),
                         labels = c("V-shaped", "Inline")),
                   width = unit(0.5,"cm"))

ht_manual<-Heatmap(mtcars[, c(9)], name = "Transmission",
                   col = colorRamp2(c(0,1),c("white", "black")),
                   heatmap_legend_param = list(at = c(0,1),
                         labels = c("Automatic", "Manual")),
                   width = unit(0.5,"cm"))
```


```{r}
ht_main = Heatmap(data.scale, name = "cluster rows",
                  cluster_rows=row.clus, cluster_columns = col.clus)

# draw() command used to put together the heatmap and the annotation 
draw(ht_main+ht_manual+ht_inline, auto_adjust = FALSE) 
```

Same HCA heatmap as above but without row labels: 
```{r}
draw(ht_main+ht_manual+ht_inline) 
```


## Part II: 

#### Select at least three different options/alternatives from ComplexHeatmap's first three chapters, and apply them to the mtcars dataset. For example, such as using different clustering algorithms (k-means, etc.), separation of heatmaps, stacking, different ways to annotate (columns and rows and different graphing options beyond black/white blocks), etc. Please provide some text in your markdown discussing what options/alternatives you selected, why this was interesting, and any challenges you experienced with running the visualizations. Provide a few sentences to interpret each new heatmap visualization.





mtcars data set 


```{r}
mtcars 
data_mat <- as.matrix(mtcars) 
```

Custom HCA heatmap 1 [pick up here]


```{r}
ha = HeatmapAnnotation(foo = 1:10) 
column_ha = HeatmapAnnotation(foo1 = runif(10), bar1 = anno_barplot(runif(10)))

Heatmap(data_mat, row_km = 2, top_annotation = column_ha) 

mtcars

# set cyl as annotation on heatmap 

ht_cyl <-Heatmap(mtcars[, c(2)], name = "Cylinder Count",
                   col = colorRamp2(c(4,6,8),c("white", "blue", "black")),
                   heatmap_legend_param = list(at = c(0,1),
                         labels = c("4", "6", "8")),
                   width = unit(0.5,"cm"))

ht_main <- Heatmap(data_mat, name = "cluster rows", 
                   cluster_rows = row.clus, cluster_columns = col.clus)


draw(ht_main + ht_cyl, auto_adjust = FALSE) 


```







