---
title: "Exercise 4"
author: "Nilima Ajaikumar, Hagay Jalon & Gian Zlupko"
date: "2022-11-14"
output: html_document
---

```{r message = FALSE, warning = FALSE}
library(tidyverse) 
library(xlsx) 
library(pROC) 
library(OptimalCutpoints) # for dealing with cutoff point
library(naniar) # replaces missing data to NA
```


# Part I: NECS Data 

#### Import Data 

The following steps were implemented to access data from NCES' online repository. For our submission, we selected a wide range of predictors in addition to some of the predictor and outcome used in Bowers & Zhou (2019). The code below was used to load and clean the raw NCES data set. To cast a wide of potential predictor variables, we use a tidyverse approach with the `select()` function, specifically using a `var:var:` call within `select()` to capture a relatively large subset of rows. 
```{r}
setwd("/Users/gianzlupko/Desktop/ORLA 6541 Data Science/ORLA_6541_Applied_Data_Science/NCES_Data") 

# Load R Data File
load("els_full_data.rdata")

# simplify name of data frame 
els_full_data <- els_02_12_byf3pststu_v1_0

# outcome variables to add - drop out: F2EVERDO, College Enrollment: F2PS0601, GPA: F1RGPP2, post-secondary STEM: F3TZSTEM1TOT
# add these outcome variables as well as some of the predictors used in Bowers and Zhou (2019). To select a wider rang e
els_clean <- els_full_data %>%
   select(STU_ID:bytxmthn, F2PS0601, F2EVERDO, F1RGPP2, F3TZSTEM1TOT) %>% rename(drop_out = F2EVERDO, college_enrollment = F2PS0601, GPA = F1RGPP2, post_sec_stem = F3TZSTEM1TOT)
```

