---
title: "Exercise 3 - HCA & Heatmaps"
author: "Nilima Ajaikumar, Hagay Jalon, and Gian Zlupko"
date: '2022-10-24'
output: html_document
---


# Exercise 3
## Hierarchical Cluster Analysis and Heatmaps

## Part I: 
#### Please replicate the code and markdown provided in the readings (Bowers) for the mtcars dataset to visualize it using hierarchical cluster analysis heatmaps following the recommendations from the readings, including using uncentered correlation.


Libraries used throughout exercise 
```{r, warning = FALSE, message = FALSE }
library(ggplot2)
library(tidyverse) 
library(broom)
library(data.table) 
library(ComplexHeatmap)
library(circlize)
library(hopach)
```

Load data 
```{r}
# use first seven variables in mtcars data set 
data<-mtcars[, c(1:7)]

# scale the data set 
data.scale<-scale(data)

# inspect the data set 
head(data.scale)
```

Generate initial heatmaps 
```{r}
# heatmap with no clustering
Heatmap(data.scale, name = "mtcars heatmap", 
        cluster_rows=FALSE , cluster_columns = FALSE)

# now cluster by rows but not columns 
Heatmap(data.scale, name = "mtcars heatmap", 
        cluster_columns = FALSE)
```


```{r}
# cluster by rows and columns 
Heatmap(data.scale, name = "mtcars heatmap")
```

Generate custom function to calculate uncentered correlation

The function creates a distance matrix, which is accessed in the `hopach` library. Per the package documentation, distance matrices contain pair wise distances between all rows of the matrix.

```{r}
uncenter.dist<-function(m) {
  as.dist(as.matrix(distancematrix(m, d="cosangle")))
}

```



```{r}
# apply function to create distance matrix on the data set
row.clus<-hclust(uncenter.dist(data.scale), method = "ave")
col.clus<-hclust(uncenter.dist(t(data.scale)), method = "ave")

# apply uncentered correlation as clustering technique and average linkage as agglomeration method 
Heatmap(data.scale, name = "mtcars heatmap", 
        cluster_rows=row.clus, cluster_columns=col.clus)

```


Heatmaps with dichotomous variables 
```{r}

ht_inline<-Heatmap(mtcars[, c(8)], name = "Cylinder Config",
                   col = colorRamp2(c(0,1),c("white", "black")),
                   heatmap_legend_param = list(at = c(0,1),
                         labels = c("V-shaped", "Inline")),
                   width = unit(0.5,"cm"))

ht_manual<-Heatmap(mtcars[, c(9)], name = "Transmission",
                   col = colorRamp2(c(0,1),c("white", "black")),
                   heatmap_legend_param = list(at = c(0,1),
                         labels = c("Automatic", "Manual")),
                   width = unit(0.5,"cm"))
```


```{r}
ht_main = Heatmap(data.scale, name = "cluster rows",
                  cluster_rows=row.clus, cluster_columns = col.clus)

# draw() command used to put together the heatmap and the annotation 
draw(ht_main+ht_manual+ht_inline, auto_adjust = FALSE) 
```

Same HCA heatmap as above but without row labels: 
```{r}
draw(ht_main+ht_manual+ht_inline) 
```


## Part II: 

#### Select at least three different options/alternatives from ComplexHeatmap's first three chapters, and apply them to the mtcars dataset. For example, such as using different clustering algorithms (k-means, etc.), separation of heatmaps, stacking, different ways to annotate (columns and rows and different graphing options beyond black/white blocks), etc. Please provide some text in your markdown discussing what options/alternatives you selected, why this was interesting, and any challenges you experienced with running the visualizations. Provide a few sentences to interpret each new heatmap visualization.




mtcars data set 


```{r}
# create new data set 
data_mat <- as.matrix(mtcars) 

# scale data 
data_mat <- scale(data_mat) 
```

Custom HCA heatmap 1 - 

Visualize a heatmap using a categorical 


```{r}

# update row and col clustering with new data set input into the custom functions created 

row.clus<-hclust(uncenter.dist(data_mat), method = "ave")
col.clus<-hclust(uncenter.dist(t(data_mat)), method = "ave")

# set cyl as annotation on heatmap 

ht_cyl <-Heatmap(mtcars[, c(2)], name = "Cylinder Count",
                   col = colorRamp2(c(4,6,8),c("grey", "blue", "black")),
                   heatmap_legend_param = list(at = c(0,1,2),
                         labels = c("4", "6", "8")),
                   width = unit(0.5,"cm"))

ht_main <- Heatmap(data_mat, name = "cluster rows", 
                   cluster_rows = row.clus, cluster_columns = col.clus)

# put together annotated heatmap using draw() function 
draw(ht_main + ht_cyl, auto_adjust = FALSE) 

```
 
 For the HCA heatmap above, we visualized engine cylinder count as a categorical variable for annotation. We ran into challenges updating the given code to annotate heatmaps. Specifically, we received the error: "The length of 'at' should be the same as the length of 'labels'. Initially, this error message appeared somewhat cryptic to us. However, upon looking at the code chunk used to edit the data corresponding to heatmap annotation, we realized that we needed to pass a third value through the 'at' command in the that the length of our variable needed to match the `heatmap_legend_param` argument. After adding a third value, corresponding with the three levels of our categorical variable, we were then able to successfully render the visualization.






