---
title: "ORLA_6541_midterm"
author: "Gian Zlupko"
date: "2022-11-01"
output: html_document
---


# ORLA 6541 Midterm 

The current Rmarkdown file contains code and supporting documentation for the following study titled, *Individual Traits and Behaviors Related to Responsibility-Taking Behaviors (RTB): An Application of Hierarchical Cluster Analysis and Heatmapping*. This research explores the relationship between individual differences, motivation, and key employee behaviors and RTB. The paper write up provides an in-depth review of the RTB construct and its validation to date. Building on previous validation research, the current study seeks to better understand individual traits and behaviors that relate to RTB. 


#### Libraries and Set Up 

```{r, warning = FALSE, message = FALSE}
library(tidyverse) 
library(ComplexHeatmap)
library(xlsx) 
library(psych) 

# working directory 
setwd("/Users/gianzlupko/Desktop/ORLA 6541 Data Science/ORLA_6541_Applied_Data_Science/Midterm")

# load data 

study_data <- read.xlsx("rtb_study1_composites.xlsx", sheetIndex = 1)

# inspect data
str(study_data) 

# data cleaning - renaming and drop second conscientiousness measure 

study_data <- study_data %>%
  rename(accountability = accn_composite, furtherance = furth_composite, ownership = own_composite, encouragement = enc_composite, RTB_score = RTB_total., 
         conscientiousness = conscientiousness_1, OCB = org_citizenship_behavior) %>% 
  select(-conscientiousness_2) 

```


#### Descriptive Statistics 

```{r}
# calculate descriptive statistics 
descriptives <- describe(study_data) 

# inspect descriptives 
descriptives

# clean descriptives data
desc_short <- descriptives %>% 
  select(mean, median, sd, min, max, skew, kurtosis)

# inspect descriptives 
desc_short 

# export descriptives to local for formatting 
#write.xlsx(desc_short, "descriptive_stats_midterm.xlsx") 

```



## Exploratory Data Visualization  

#### Viz I: Relationship between RTB, Conscientiousness, and Employee Engagement 

H1a: The relationship between RTB and conscientiousness will be moderated by employee engagement. Specifically, when conscientiousness is high and EE is high, RTB will be high. In contrast, when conscientiousness is high but EE is low, RTB will be lower. 

To build this visualization, I first collapsed EE to an ordinal variable to denote low, meidum, and high levels of EE. I store this as a factor data type, which will come in handy in the following split line plot, where 3 separte regression lines are visualized corresponding with the level of EE. 
```{r}

# create new var for new levels of EE 
ee_data <- study_data %>% 
  mutate(ee_levels = as.factor(ifelse(0 <=employee_engagement & employee_engagement < 1.666667, "low", ifelse(1.66667 <= employee_engagement & employee_engagement < 3.3333, "medium", "high"))))    

# show count - findings indicate that there is too little variation
ee_data %>% 
  count(ee_levels) 

# instead, I will sum to only two levels for EE - neutral and high 
# the median is 4.08 and the SD is .65 so as it is the data is left skewed with not a lot of variation. Moreover, there are very few low values of EE so the categories of 'neutral' and 'high' for EE is an accurate representation of the data

study_data %>% 
  ggplot(aes(x =employee_engagement)) + 
  geom_density(fill = "#69b3a2", alpha = 0.8)   # this density plot shows the left skew 

# create new levels of EE for a factor variable that will then be used to group the employees and show their separate regression lines. Write over previous object. 
ee_data <- study_data %>% 
  mutate(ee_levels = 
           as.factor(ifelse(0 <=employee_engagement & employee_engagement < 4, "Neutral", "High")))

# review the levels. Note, I did not split the original EE continuous variable down the middle as the data was strongly left skewed. Thus, to strive to ensure more equal subgroup sample sizes, my cut point for 'neutrally engaged' employees was 4. In reality this is typically viewed as positive engagement/motivation/absorption in one's work, but using this outpoint will make for a more effective data viz to explore the potential moderating effect of EE on the RTB and conscientiousness. The underlying regression itself will nonetheless be true to the data, effect size, significance, etc. 
ee_data %>% count(ee_levels) 
```

Build the visualization for the moderation analysis 

```{r, message = FALSE}
library(ggtext) 

ee_data %>%
  ggplot(aes(x = conscientiousness, y = RTB_score, color = ee_levels)) + 
  geom_point(position = "jitter", alpha = 0.5) + geom_smooth(method = lm, se = FALSE) +  scale_color_manual(
    values = c(High = "#F8766D", Neutral = "#619CFF"), 
    guide = "none") + labs(x = "Conscientiousness", 
                           y = "Responsibility-Taking Behavior", 
                           title ="<span style = 'font-size:16pt; font-family:Helvetica;'> Conscientious employees that were more<strong><span style = 'color:#F8766D'><b> engaged  </strong>possessed higher RTB") + 
  theme(plot.title = element_markdown(), axis.text.x=element_text(size=12), 
        axis.text.y=element_text(size = 12)) + theme(axis.title = element_text(size = 14)) 

# confirm the interaction that we see with regression analysis
# results indicate that the interaction b/w EE and conscientiousness is significant (p < .05) 
lm(RTB_score ~ conscientiousness + employee_engagement + conscientiousness*employee_engagement, 
   data = study_data) %>% summary() 

```



































